---
title: "Load Data"
format:
  html:
    embed-resources: true
editor: visual
execute:
  warning: FALSE
  message: FALSE
---

The following script will load the raw data set that contains 3 sheets to be joined:

-   The read counts for each experiment

-   The Reads Per Million (RPM) for each experiment.

-   The Reads Per Kilobase per Million mapped reads (RPKMs) for each experiment.

-   The output of this script will be a data set with all 3 sheets combined.

0.  Load Libraries

```{r}
#| output: false
library("GEOquery")
library("tidyverse")
library(readxl)
library(R.utils)
```

1.  Extract GEO dataset using the GEOquery package

```{r}
gset <- getGEO("GSE239956", GSEMatrix = TRUE, getGPL = FALSE) |> 
  pluck(1)

# gset is an object. Inspecting it, we see that our data is on a list called "other"
#Since gset is an object, we have to use slot function to extract information individual slots in an object
data_loc <- gset |> 
  experimentData() |> 
  slot("other") |> 
  pluck("supplementary_file")

# Print the URL
print(data_loc)
```

2.  Create required directories to load the data

```{r}
#| output: false
raw_dir <- "../_raw"
data_dir <- "../data"
results_dir <- "../results"
compressed_file <- str_c(raw_dir, "/raw_data.xls.gz")
decompressed_file <- str_c(raw_dir, "/raw_data.xls")

# Create required directories if they don't exist
if (!dir.exists(raw_dir)) dir.create(raw_dir)
if (!dir.exists(data_dir)) dir.create(data_dir)
if (!dir.exists(results_dir)) dir.create(results_dir)

```

3.  Load and decompress the data file

```{r}
#| output: false
if (!file.exists(decompressed_file)) {
  download.file(
    url = data_loc,
    destfile = compressed_file,
    mode = "wb"
  )
}

# Decompress the file, 
if (!file.exists(decompressed_file)) {
  gunzip(compressed_file, destname = decompressed_file, remove = FALSE)
}
```

4.  Load all available sheets

```{r}
#| output: false
reads_df <- read_excel(decompressed_file, sheet = 1)
RPMs_df <- read_excel(decompressed_file, sheet = 2)
RPKMs_df <- read_excel(decompressed_file, sheet = 3)
```

5.  Save sheets

```{r}
#| output: false
reads_df |> 
  write_tsv("../data/01_data_load_reads.tsv.gz")
RPMs_df |> 
  write_tsv("../data/01_data_load_RPMs.tsv.gz")
RPKMs_df |> 
  write_tsv("../data/01_data_load_RPKMs.tsv.gz")
```
