---
title: "Augment data"
format:
  html:
    embed-resources: true
editor: visual
---

1.  Load data and relevant libraries

```{r}
#| message: false
library("tidyverse")
library("dplyr")

```

```{r}
data_dir <- "../data"
data_file <- "/02_data_load.tsv.gz" 
aug_input <- read_tsv(str_c(data_dir, data_file))

```

2.  Normalizing the data by log transformation. Adding a small value before log transforming data, to ensure that zeros do not appear.

```{r}
df_aug <- aug_input |> 
  mutate(across(starts_with("reads") | starts_with("RPM") | starts_with("RPKM"),~ log(.+0.0001)))
```

3.  Calculating mean of the three groups, and adding the columns to the file

```{r}
data_augmented <- df_aug |> 
  mutate(
    mean_reads = rowMeans(across(starts_with("reads")), na.rm = TRUE),  
    mean_rpm = rowMeans(across(starts_with("RPM")), na.rm = TRUE),     
    mean_rpkm = rowMeans(across(starts_with("RPKM")), na.rm = TRUE)    
  )
```

4.  Our dataset is a wide dataset however, It might be of interest having our df in a long format

```{r}
data_long_augmented <- data_augmented |> 
  pivot_longer(
    cols = contains(c("control", "tau")), 
    names_to = "experiment", 
    values_to = "gene_expression"
  )
```

The data is now normalized, mean is calcualted for each category and the data is in long format.

5.  Data is saved in file:

```{r}
data_long_augmented |> 
  write_tsv(file = "../data/03_data_load.tsv.gz")

```
